---
title: "Spatial and Seasonal Distributions and Characteristics of Microplastics in Narragansett Bay Surface Water"
author: "Sarah Davis"
date: "2023-05-01"
output: html_document
---

```{r echo=TRUE}
```

***The first portion of code utilizes the data set "Master Plastics Analysis.csv" in order to:***

1)  ***Calculate and visualize how many microplastic (MP) particles on average were recovered from each sampling event and during each season (normalized for the amount of water sampled). Box plots are used to show averages and variation.***

2)  ***Explore the characteristics (morphology, color, size, and polymer type) of recovered plastic particles, particularly whether there are observable patterns in spatial or temporal variables using data visualizations. Bar graphs are used to show proportions of each characteristic out of the entire observed population of plastic particles.***

***Note that figure and table numbers refer to their order in this R-markdown document.***

```{r echo=TRUE, message=FALSE}

library(tidyverse)
library(dplyr)
library(magrittr)

#load in plastic particle data from csv
all_plastic <- read.csv("Master Plastics Analysis.csv")

#Filter the data to keep only the relevant environmental samples (no control samples)
net_samples <- filter(all_plastic, SampleType == "NS")

#summarize the number of particles per sample event, normalized for the amount of water sampled
particles_per_event <- net_samples %>%
  group_by(Date, SampleEvent, Coord_Starting_X, Coord_Starting_Y, Total_Water_Sampled_m3, Season, Site) %>%
  summarise(total_particles_sampleevent = n())

particles_per_event <- particles_per_event %>%
  mutate(Particles_per_meter_cubed = total_particles_sampleevent/Total_Water_Sampled_m3)

particles_per_event$Date <- as.Date(particles_per_event$Date, format = "%m/%d/%Y")

#rename the sites according to their numerical label for map comparison
particles_per_event$Site <- gsub("BLS", "6",
                                       gsub("CPO", "1",
                                            gsub("HOP", "3",
                                                 gsub("GIO", "4",
                                                      gsub("WDI", "5",
                                                           gsub("HOG", "2", particles_per_event$Site))))))


#calculate the average particles/m3 across both years at each site, while considering the standard error.
particles_per_site_season <- particles_per_event %>%
  group_by(Season, Site) %>%
  summarise(mean = mean(Particles_per_meter_cubed), 
            mean_se = mean_se(Particles_per_meter_cubed))

library(ggplot2)
#plot the average outcomes for each season by site in a box plot to show variation
box_particles_per_site_season_plot <- ggplot(particles_per_event, aes(x=Site, y=Particles_per_meter_cubed, fill=Season)) +
  geom_boxplot(width=0.5, coef=2, position=position_dodge(0.9)) +
  labs(x="Site", y="Average MPs per m3 of Seawater", fill="Season", title=("Average MP Concentrations across Locations and Seasons in Narragansett Bay")) +
  scale_fill_manual(values = c("#FFCA99", "#CC5800", "#1E8E99")) +
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text = element_text(color = "black"),
        axis.title = element_text(color = "black"),
        plot.title = element_blank(),
        panel.spacing = unit(0.1, "lines"))


plot(box_particles_per_site_season_plot)

```

**Figure 1. Average concentration of MPs observed at NB sampling sites across the Spring, Summer/Fall, and Winter seasons.** When considering seasonal variation, particle concentrations were consistently highest in the northern portion of NB throughout all seasons (Sites 1, 2, and 3), with high variability across seasons in the southern portion of NB (Sites 4, 5, 6). Site 5 exhibited a major spike in MP yield during the Winter season that was not observed consistently at other sampling sites. These findings highlight the spatial and seasonal dynamics of pollution in Narragansett Bay, emphasizing the importance of improved targeted monitoring and management strategies in mitigating MP pollution.

```{r echo=TRUE, message=FALSE}

#summarize and visualize particle morphology data by season
morphology_summary_season_table <- net_samples %>%
  filter(!is.na(Shape.Category)) %>%
  group_by(Season) %>%
  summarise(
    Fibers = sum(Shape.Category == "Fiber"),
    Fragments = sum(Shape.Category == "Fragment"),
    Beads = sum(Shape.Category == "Bead")
  )

morphology_summary_season <- morphology_summary_season_table %>%
  transform(Total = Fibers + Fragments + Beads)

morphology_summary_season$Fibers <- morphology_summary_season$Fibers / morphology_summary_season$Total * 100
morphology_summary_season$Fragments <- morphology_summary_season$Fragments / morphology_summary_season$Total * 100
morphology_summary_season$Beads <- morphology_summary_season$Beads / morphology_summary_season$Total * 100

#summarize and visualize morphology data by site
morphology_summary_site_table <- net_samples %>%
  filter(!is.na(Shape.Category)) %>%
  group_by(Site) %>%
  summarise(
    Fibers = sum(Shape.Category == "Fiber"),
    Fragments = sum(Shape.Category == "Fragment"),
    Beads = sum(Shape.Category == "Bead")
  )
morphology_summary_site <- morphology_summary_site_table %>%
  transform(Total = Fibers + Beads + Fragments)

morphology_summary_site$Fibers <- morphology_summary_site$Fibers / morphology_summary_site$Total * 100
morphology_summary_site$Fragments <- morphology_summary_site$Fragments / morphology_summary_site$Total * 100
morphology_summary_site$Beads <- morphology_summary_site$Beads / morphology_summary_site$Total * 100

morphology_summary_site$Site <- gsub("BLS", "1",
                                     gsub("CPO", "4",
                                          gsub("HOP", "5",
                                               gsub("GIO", "2",
                                                    gsub("WDI", "6",
                                                         gsub("HOG", "3", morphology_summary_site$Site))))))



#Plot morphology by season
morphology_season_long <- morphology_summary_season %>% 
  select(-Total) %>%
  gather(key = "category", value = "count", -Season)

morphology_season_plot <- ggplot(morphology_season_long, aes(x = Season, y = count, fill = category)) +
  geom_bar(stat = "identity") +
  labs(x = "Season", y = "Percent of Total MPs", fill = "Particle Type", title="MP Morphology by Season in Narragansett Bay") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(panel.background = element_blank()) +
  scale_fill_manual(values = c("#8F8782", "#A5ACAF", "#414451")) +
  theme(axis.line = element_line(color = "black", linewidth = 0.5),
        axis.title = element_text(color = "black"))

#Plot morphology by site
morphology_site_long <- morphology_summary_site %>%
  select(-Total) %>%
  gather(key = "category", value = "count", -Site)

morphology_site_plot <- ggplot(morphology_site_long, aes(x=Site, y = count, fill = category)) +
  geom_bar(stat = "identity") +
  labs(x = "Site", y = "Percent of Total MPs", fill = "Particle Type", title="MP Morphology by Site in Narragansett Bay") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(panel.background = element_blank()) +
  scale_fill_manual(values = c("#8F8782", "#A5ACAF", "#414451")) +
  theme(axis.line = element_line(color = "black", linewidth = 0.5),
        axis.title = element_text(color = "black"))

plot(morphology_season_plot)
plot(morphology_site_plot)

```

**Figures 2 and 3. Morphology of MPs across seasons and sites.** Morphology was not significantly variable across seasons or sites sampled, with fragments comprising the majority of particles recovered. Fibers and beads were also present but in relatively lower quantities. These results suggest that the assemblage of microplastics in Narragansett Bay remains consistent across seasons and sites, with fragmented particles being the dominant form. Understanding the morphology of MPs provides insights into potential sources and pathways of pollution in the bay. \

```{r echo=TRUE, message=FALSE}

#summarize and visualize color by site
color_summary_site <- net_samples %>%
  filter(!is.na(Color)) %>%
  group_by(Site) %>%
  summarise(
    Clear = sum(Color == "Clear"),
    White = sum(Color == "White"),
    Black = sum(Color == "Black"),
    Blue = sum(Color == "Blue"),
    Yellow = sum(Color == "Yellow"),
    Red = sum(Color == "Red"),
    Orange = sum(Color == "Orange"),
    Tan = sum(Color == "Tan/Beige"),
    Pink = sum(Color == "Pink"),
    Violet = sum(Color == "Violet"),
    Brown = sum(Color == "Brown"),
    Grey = sum(Color == "Grey"),
    Green = sum(Color == "Green")
  ) %>%
  transform(Total = Clear + White + Black + Blue + Yellow + Red + Orange + Tan + Pink + Violet + Brown + Grey + Green)

color_summary_site$Clear <- color_summary_site$Clear / color_summary_site$Total * 100
color_summary_site$White <- color_summary_site$White / color_summary_site$Total * 100
color_summary_site$Black <- color_summary_site$Black / color_summary_site$Total * 100
color_summary_site$Blue <- color_summary_site$Blue / color_summary_site$Total * 100
color_summary_site$Yellow <- color_summary_site$Yellow / color_summary_site$Total * 100
color_summary_site$Red <- color_summary_site$Red / color_summary_site$Total * 100
color_summary_site$Orange <- color_summary_site$Orange / color_summary_site$Total * 100
color_summary_site$Tan <- color_summary_site$Tan / color_summary_site$Total * 100
color_summary_site$Pink <- color_summary_site$Pink / color_summary_site$Total * 100
color_summary_site$Violet <- color_summary_site$Violet / color_summary_site$Total * 100
color_summary_site$Brown <- color_summary_site$Brown / color_summary_site$Total * 100
color_summary_site$Grey <- color_summary_site$Grey / color_summary_site$Total * 100
color_summary_site$Green <- color_summary_site$Green / color_summary_site$Total * 100

color_site_long <- color_summary_site %>%
  select(-Total) %>%
  gather(key = "category", value = "count", -Site)

color_site_long$Site <- gsub("BLS", "1",
                               gsub("CPO", "4",
                                    gsub("HOP", "5",
                                         gsub("GIO", "2",
                                              gsub("WDI", "6",
                                                   gsub("HOG", "3", color_site_long$Site))))))

particle_colors <- c("Clear" = "ghostwhite", "White" = "white", "Black" = "black", "Blue" = "blue", "Violet" = "violet", "Red" = "red", "Pink" = "pink", "Orange" = "orange", "Yellow" = "yellow", "Tan" = "tan", "Brown" = "brown", "Green" = "#4CBB17", "Grey" = "grey")

color_site_plot <- ggplot(color_site_long, aes(x = Site, y = count, fill = reorder(category, count))) +
  geom_bar(stat = "identity", color = "black") +  # add black outline to bars
  labs(x = "Site", y = "Percent of Total MPs", fill = "Category", title="MP Color by Site in Narragansett Bay") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = particle_colors) +
  theme(panel.background = element_blank()) +
  theme(axis.line = element_line(color = "black", linewidth = 0.5),
        axis.title = element_text(color = "black"))


#summarize and visualize color by season 
color_summary_season <- net_samples %>%
  filter(!is.na(Color)) %>%
  group_by(Season) %>%
  summarise(
    Clear = sum(Color == "Clear"),
    White = sum(Color == "White"),
    Black = sum(Color == "Black"),
    Blue = sum(Color == "Blue"),
    Yellow = sum(Color == "Yellow"),
    Red = sum(Color == "Red"),
    Orange = sum(Color == "Orange"),
    Tan = sum(Color == "Tan"),
    Pink = sum(Color == "Pink"),
    Violet = sum(Color == "Violet"),
    Brown = sum(Color == "Brown"),
    Grey = sum(Color == "Grey"),
    Green = sum(Color == "Green")
  ) %>%
  transform(Total = Clear + White + Black + Blue + Yellow + Red + Orange + Tan + Pink + Violet + Brown + Grey + Green)

color_summary_season$Clear <- color_summary_season$Clear / color_summary_season$Total * 100
color_summary_season$White <- color_summary_season$White / color_summary_season$Total * 100
color_summary_season$Black <- color_summary_season$Black / color_summary_season$Total * 100
color_summary_season$Blue <- color_summary_season$Blue / color_summary_season$Total * 100
color_summary_season$Yellow <- color_summary_season$Yellow / color_summary_season$Total * 100
color_summary_season$Red <- color_summary_season$Red / color_summary_season$Total * 100
color_summary_season$Orange <- color_summary_season$Orange / color_summary_season$Total * 100
color_summary_season$Tan <- color_summary_season$Tan / color_summary_season$Total * 100
color_summary_season$Pink <- color_summary_season$Pink / color_summary_season$Total * 100
color_summary_season$Violet <- color_summary_season$Violet / color_summary_season$Total * 100
color_summary_season$Brown <- color_summary_season$Brown / color_summary_season$Total * 100
color_summary_season$Grey <- color_summary_season$Grey / color_summary_season$Total * 100
color_summary_season$Green <- color_summary_season$Green / color_summary_season$Total * 100

color_season_long <- color_summary_season %>%
  select(-Total) %>%
  gather(key = "category", value = "count", -Season)

particle_colors <- c("Clear" = "ghostwhite", "White" = "white", "Black" = "black", "Blue" = "blue", "Violet" = "violet", "Red" = "red", "Pink" = "pink", "Orange" = "orange", "Yellow" = "yellow", "Tan" = "tan", "Brown" = "brown", "Green" = "#4CBB17", "Grey" = "grey")

color_season_plot <- ggplot(color_season_long, aes(x = Season, y = count, fill = reorder(category, count))) +
  geom_bar(stat = "identity", color = "black") +  # add black outline to bars
  labs(x = "Season", y = "Percent of Total MPs", fill = "Category", title="MP Color by Season in Narragansett Bay") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = particle_colors) +
  theme(panel.background = element_blank()) +
  theme(axis.line = element_line(color = "black", linewidth = 0.5),
        axis.title = element_text(color = "black"))

plot(color_season_plot)
plot(color_site_plot)
```

**Figures 4 and 5. Proportion of MP particle colors observed across seasons and sites.**  A diverse range of particle colors was identified, with white, clear, blue, and black being the dominant colors, collectively comprising a significant proportion of the observed MPs. Importantly, the proportions of different colors did not show significant variation across different sites or seasons, suggesting a consistent color distribution pattern in the plastics present in Narragansett Bay.

```{r echo=TRUE, message=FALSE}

#summarize particle length by site and season
avg_length_site_season <- net_samples %>%
  filter(!is.na(L..um.)) %>%
  filter(!is.na(W..um.)) %>%
  group_by(Site, Season) %>%
  summarise(avg_length = mean(L..um.), std_error = mean_se(L..um.))
  
avg_length_site_season$Site <- gsub("BLS", "1",
                                    gsub("CPO", "4",
                                         gsub("HOP", "5",
                                              gsub("GIO", "2",
                                                   gsub("WDI", "6",
                                                        gsub("HOG", "3", avg_length_site_season$Site))))))
#visualize particle length averages
avg_length_plot <- ggplot(avg_length_site_season, aes(x=Site, y = avg_length, fill = Season)) +
  geom_bar(stat="identity", position=position_dodge()) +
  geom_errorbar(aes(ymin=std_error$ymin, ymax=std_error$ymax), width=.2, position=position_dodge(.9)) +
  labs(x="Site", y="Average Particle Length (μm)", fill="Season", title="Average MP Length by Site and Season in Narragansett Bay") +
  scale_fill_manual(values = c("#FFCA99", "#CC5800", "#1E8E99")) +
  theme(panel.background = element_blank()) +
  theme(axis.line = element_line(color = "black", linewidth = 0.5),
        axis.title = element_text(color = "black"))

plot(avg_length_plot)
```

**Figure 6.** **Average MP length at each site by season.** In Spring, Sites 2, 3, 4, and 5 displayed the shortest particle length, indicating a prevalence of smaller-sized microplastics during this season. However, in Winter, these same sites exhibited the longest particle length, suggesting an increase in the proportion of larger-sized MPs. These findings suggest that seasonal dynamics are a stronger driver of differences in particle length than site location.

```{r echo=TRUE, message=FALSE}

#summarize and visualize polymer data by season
polymer_season <- net_samples %>%
  filter(!is.na(material_ID)) %>%
  group_by(Season) %>%
  summarise(
    PE = sum(material_ID == "PE"),
    PP = sum(material_ID =="PP"),
    Nylon = sum(material_ID == "Nylon"),
    PVC = sum(material_ID == "PVC"),
    EVA = sum(material_ID == "EVA"),
    PS = sum(material_ID == "PS"), 
    Other = sum(material_ID %in% c("ABS", "Acrylic", "CR", "Diisooctyl Adepate", "Ethylene Propylene Diene Terpolymer", "PA", "PAN", "PC", "PET", "PET.1", "PEVA", "Polyester", "Polyisoprene", "PVA"))
  ) %>%
  transform(Total = PE + PP + Nylon + EVA + PVC + PS + Other)


#calculate proportions of total for each polymer type
polymer_season$PP <- polymer_season$PP / polymer_season$Total * 100
polymer_season$PE <- polymer_season$PE / polymer_season$Total * 100
polymer_season$Nylon <- polymer_season$Nylon / polymer_season$Total * 100
polymer_season$PVC <- polymer_season$PVC / polymer_season$Total * 100
polymer_season$EVA <- polymer_season$EVA / polymer_season$Total * 100
polymer_season$PS <- polymer_season$PS / polymer_season$Total * 100
polymer_season$Other <- polymer_season$Other / polymer_season$Total *100

#make the data long
polymer_season_long <- polymer_season %>%
  select(-Total) %>%
  gather(key = "category", value = "count", -Season)

library(paletteer)
polymer_palette <- paletteer_d("ggsci::default_locuszoom")
                               

polymer_season_plot <- ggplot(polymer_season_long, aes(x = Season, y = count, fill = reorder(category, count))) +
  geom_bar(stat = "identity") +
  labs(x = "Season", y = "Percent of Total MPs", fill = "Polymer Type", title="MP Polymer Type by Season in Narragansett Bay") +
  scale_fill_manual(values = polymer_palette) +
  theme(panel.background = element_blank()) +
  theme(axis.line = element_line(color = "black", linewidth = 0.5),
        axis.title = element_text(color = "black"))

#summarize and visualize polymer data by site
polymer_site <- net_samples %>%
  filter(!is.na(material_ID)) %>%
  group_by(Site) %>%
  summarise(
    PE = sum(material_ID == "PE"),
    PP = sum(material_ID =="PP"),
    Nylon = sum(material_ID == "Nylon"),
    PVC = sum(material_ID == "PVC"),
    EVA = sum(material_ID == "EVA"),
    PS = sum(material_ID == "PS"), 
    Other = sum(material_ID %in% c("ABS", "Acrylic", "CR", "Diisooctyl Adepate", "Ethylene Propylene Diene Terpolymer", "PA", "PAN", "PC", "PET", "PET.1", "PEVA", "Polyester", "Polyisoprene", "PVA"))
  ) %>%
  transform(Total = PE + PP + Nylon + EVA + PVC + PS + Other)

#calculate proportions of total for each polymer type
polymer_site$PP <- polymer_site$PP / polymer_site$Total * 100
polymer_site$PE <- polymer_site$PE / polymer_site$Total * 100
polymer_site$Nylon <- polymer_site$Nylon / polymer_site$Total * 100
polymer_site$PVC <- polymer_site$PVC / polymer_site$Total * 100
polymer_site$EVA <- polymer_site$EVA / polymer_site$Total * 100
polymer_site$PS <- polymer_site$PS / polymer_site$Total * 100
polymer_site$Other <- polymer_site$Other / polymer_site$Total *100

#make the data long
polymer_site_long <- polymer_site %>%
  select(-Total) %>%
  gather(key = "category", value = "count", -Site)

#rename the sites to numerical for map comparison
polymer_site_long$Site <- gsub("BLS", "6",
                                       gsub("CPO", "1",
                                            gsub("HOP", "3",
                                                 gsub("GIO", "4",
                                                      gsub("WDI", "5",
                                                           gsub("HOG", "2", polymer_site_long$Site))))))

polymer_site_plot <- ggplot(polymer_site_long, aes(x=Site, y = count, fill = reorder(category, count))) +
  geom_bar(stat = "identity") +
  labs(x = "Site", y = "Percent of Total MPs", fill = "Polymer Type", title="MP Polymer Type by Site in Narragansett Bay") +
  scale_fill_manual(values = polymer_palette) +
  theme(panel.background = element_blank()) +
  theme(axis.line = element_line(color = "black", linewidth = 0.5),
        axis.title = element_text(color = "black"))

plot(polymer_season_plot)
plot(polymer_site_plot)
```

**Figures 7 and 8. Proportions of identified MP polymer types by season and site.** Polyethylene (PE) and polypropylene (PP) were found to be the predominant polymer types, comprising over 87% of all recovered MPs. The proportions of polymer types did not exhibit significant variation among different sites or seasons. It is important to acknowledge that the surface trawl sampling method used in this study is more likely to capture low-density floating particles, which may explain the higher representation of PE, PP, and polystyrene (PS) in the recovered samples.

***The second portion of code utilizes the data set "sample_event_data.csv" in order to:***

1)  ***Model and visualize the linear relationship between latitude and MP particle yield.***\

```{r, echo=TRUE, message=FALSE}

sampling_event_data <- read.csv("sample_event_data.csv")

#model linear relationship between latitude (significant predictor variable) and MP particle yield
latitude_lm <- lm(data=sampling_event_data, Particles_per_meter_cubed~Coord_Starting_Y)

stats <- summary(latitude_lm)
print(stats)

#create a scatter plot of the linear model with the regression line
lat <- ggplot(sampling_event_data, aes(x = Coord_Starting_Y, y = Particles_per_meter_cubed)) +
  geom_point(color = "blue", size = 3) +
  geom_smooth(method = "lm", formula = y ~ x, color = "red", se = FALSE) +
  labs(x = "Latitude", y = "MP Yields",
       title = "Latitude vs. Observed MP Concentrations in Narragansett Bay") +
  theme_minimal() +
  theme(panel.grid = element_blank(), 
        panel.background = element_blank(),
        axis.line = element_line(color = "black"))

plot(lat)


```

**Figure 9. Relationship between latitude and MP concentrations in Narragansett Bay surface water.** Scatter plot showing the distribution of particle yields (Particles/meter3) across different latitudes in Narragansett Bay. The red regression line represents the linear relationship between latitude and particle yields.. Blue data points represent individual sample event observations. The plot demonstrates a significant positive relationship (p = 0.00927), indicating that particle yields tend to increase as latitude increases in Narragansett Bay. The plot provides insights into the spatial distribution of MPs in the bay and highlights the potential influence of predictors (freshwater inputs, population density, stormwater runoff) on MP pollution patterns.
